\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2019-12-11 analysis/paper/paper.tex

\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \DeclareUnicodeCharacter{2500}{-}

@@

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were further reduced
through manual curation to a subset of 14 packages. These packages
appears in Table \ref{tab:surv-pkgs}, along with some of the
functionalities they offer.
\textcolor{red}{JESSE: It's not clear how you selected this subset, going from 60 to 14 is a big drop.}

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: \pkg{CoxRidge}
\citeyearpar{perperoglou}, \pkg{rstpm2} \citeyearpar{clements_liu} and
\pkg{survival} \citeyearpar{survival-package}. On the other hand,
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, this is the only package to offer
this functionality.
\textcolor{red}{JESSE: Is this true? Because it seems to be a logical conclusion from your paragraph.}

@@

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
R+                        levels = c(0, 1), 
R+                        labels = c("Control group", "Screening group"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
R> plot(pt_object)
\end{CodeInput}

\begin{center}\includegraphics{../figures/plot-erspc-data-1} \end{center}

@@

\begin{CodeInput}
R> pt_object_strat <- casebase::popTime(ERSPC, 
R+                                      event = "DeadOfPrCa", 
R+                                      exposure = "ScrArm")
R> plot(pt_object_strat)
\end{CodeInput}

\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-1} \end{center}

@@

\begin{CodeInput}
R> plot(pt_object_strat, ncol = 2)
\end{CodeInput}

\begin{center}\includegraphics{../figures/plot-stratified-erspc-data-side-1} \end{center}

@@

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
R+                              data = ERSPC)
R> summary(cox_model)
\end{CodeInput}

\begin{CodeOutput}
#> Call:
#> survival::coxph(formula = Surv(Follow.Up.Time, DeadOfPrCa) ~ 
#>     ScrArm, data = ERSPC)
#> 
#>   n= 159893, number of events= 540 
#> 
#>                         coef exp(coef) se(coef)     z Pr(>|z|)  
#> ScrArmScreening group -0.222     0.801    0.088 -2.52    0.012 *
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#>                       exp(coef) exp(-coef) lower .95 upper .95
#> ScrArmScreening group     0.801       1.25     0.674     0.952
#> 
#> Concordance= 0.519  (se = 0.011 )
#> Likelihood ratio test= 6.45  on 1 df,   p=0.01
#> Wald test            = 6.37  on 1 df,   p=0.01
#> Score (logrank) test = 6.39  on 1 df,   p=0.01
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R+                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> legend("topleft", 
R+        legend = c("Control group", "Screening group"), 
R+        col = c("red","blue"),
R+        lty = c(1, 1), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-cox-cif-1} \end{center}

@@

\begin{CodeInput}
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
R+                                                   data = ERSPC, 
R+                                                   ratio = 100)
R> 
R> summary(casebase_exponential)
\end{CodeInput}

\begin{CodeOutput}
#> 
#> Call:
#> glm(formula = formula, family = binomial, data = sampleData)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.148  -0.148  -0.148  -0.133   3.079  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value            Pr(>|z|)    
#> (Intercept)            -7.7650     0.0557 -139.44 <0.0000000000000002 ***
#> ScrArmScreening group  -0.2166     0.0884   -2.45               0.014 *  
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 6052.9  on 54538  degrees of freedom
#> AIC: 6057
#> 
#> Number of Fisher Scoring iterations: 7
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_exponential)[2])
\end{CodeInput}

\begin{CodeOutput}
#> ScrArmScreening group 
#>                  0.81
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_exponential)[2,])
\end{CodeInput}

\begin{CodeOutput}
#>  2.5 % 97.5 % 
#>   0.68   0.96
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-cif-1} \end{center}

@@

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
R+                                  data = ERSPC, 
R+                                  ratio = 100)
R> 
R> summary(casebase_time)
\end{CodeInput}

\begin{CodeOutput}
#> 
#> Call:
#> glm(formula = formula, family = binomial, data = sampleData)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.389  -0.161  -0.122  -0.095   3.464  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value            Pr(>|z|)    
#> (Intercept)            -9.0257     0.1122  -80.42 <0.0000000000000002 ***
#> Follow.Up.Time          0.2207     0.0145   15.18 <0.0000000000000002 ***
#> ScrArmScreening group  -0.2311     0.0886   -2.61              0.0091 ** 
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059  on 54539  degrees of freedom
#> Residual deviance: 5815  on 54537  degrees of freedom
#> AIC: 5821
#> 
#> Number of Fisher Scoring iterations: 8
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
#>           (Intercept)        Follow.Up.Time ScrArmScreening group 
#>               0.00012               1.24701               0.79370
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_time))
\end{CodeInput}

\begin{CodeOutput}
#>                          2.5 %  97.5 %
#> (Intercept)           0.000096 0.00015
#> Follow.Up.Time        1.212086 1.28321
#> ScrArmScreening group 0.666385 0.94341
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R+                                           time = seq(0,15,0.1), 
R+                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-weibull-cif-1} \end{center}

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
R+                                     data = ERSPC, 
R+                                     ratio = 100)
R> 
R> summary(casebase_splines)
\end{CodeInput}

\begin{CodeOutput}
#> 
#> Call:
#> glm(formula = formula, family = binomial, data = sampleData)
#> 
#> Deviance Residuals: 
#>    Min      1Q  Median      3Q     Max  
#> -0.397  -0.173  -0.136  -0.085   3.797  
#> 
#> Coefficients:
#>                       Estimate Std. Error z value             Pr(>|z|)    
#> (Intercept)           -10.2787     0.3175  -32.38 < 0.0000000000000002 ***
#> bs(Follow.Up.Time)1     4.3191     0.8062    5.36        0.00000008456 ***
#> bs(Follow.Up.Time)2     2.1061     0.4878    4.32        0.00001580161 ***
#> bs(Follow.Up.Time)3     4.5310     0.7160    6.33        0.00000000025 ***
#> ScrArmScreening group  -0.2068     0.0886   -2.33                 0.02 *  
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
#> 
#> (Dispersion parameter for binomial family taken to be 1)
#> 
#>     Null deviance: 6059.0  on 54539  degrees of freedom
#> Residual deviance: 5792.2  on 54535  degrees of freedom
#> AIC: 5802
#> 
#> Number of Fisher Scoring iterations: 9
\end{CodeOutput}

\begin{CodeInput}
R> exp(coef(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
#>           (Intercept)   bs(Follow.Up.Time)1   bs(Follow.Up.Time)2 
#>              0.000034             75.119228              8.216108 
#>   bs(Follow.Up.Time)3 ScrArmScreening group 
#>             92.847641              0.813203
\end{CodeOutput}

\begin{CodeInput}
R> exp(confint(casebase_splines))
\end{CodeInput}

\begin{CodeOutput}
#>                           2.5 %     97.5 %
#> (Intercept)            0.000018   0.000062
#> bs(Follow.Up.Time)1   15.997506 378.737706
#> bs(Follow.Up.Time)2    3.256838  22.143147
#> bs(Follow.Up.Time)3   21.638298 362.098155
#> ScrArmScreening group  0.682786   0.966562
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R+                                     time = seq(0,15,0.1), 
R+                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R+      xlab = "Years since Randomization", 
R+      ylab = "Cumulative Incidence (%)", 
R+      fun = "event",
R+      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R+      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R+                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R+                     exp(coef(cox_model)), 
R+                     exp(confint(cox_model))[1], 
R+                     exp(confint(cox_model))[2]),
R+      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R+        legend = c("Control group (Cox)","Control group (Casebase)",
R+                   "Screening group (Cox)", "Screening group (Casebase)"), 
R+        col = c("red","red", "blue","blue"),
R+        lty = c(1, 2, 1, 2), 
R+        bg = "gray90")
\end{CodeInput}

\begin{center}\includegraphics{../figures/erspc-casebase-splines-cif-1} \end{center}

@@

\begin{tabular}{l|r|l|r}
\hline
Model & HR & CI & Width\\
\hline
Cox & 0.80 & (0.67, 0.95) & 0.28\\
\hline
Exponential & 0.81 & (0.68, 0.96) & 0.28\\
\hline
Weibull & 0.79 & (0.67, 0.94) & 0.28\\
\hline
Splines & 0.81 & (0.68, 0.97) & 0.28\\
\hline
\end{tabular}

@@

\begin{CodeInput}
R> anova(casebase_exponential, casebase_splines, test = "LRT")
\end{CodeInput}

\begin{CodeOutput}
#> Analysis of Deviance Table
#> 
#> Model 1: DeadOfPrCa ~ ScrArm + offset(offset)
#> Model 2: DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm + offset(offset)
#>   Resid. Df Resid. Dev Df Deviance            Pr(>Chi)    
#> 1     54538       6053                                    
#> 2     54535       5792  3      261 <0.0000000000000002 ***
#> ---
#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> AIC(casebase_exponential)
\end{CodeInput}

\begin{CodeOutput}
#> [1] 6057
\end{CodeOutput}

\begin{CodeInput}
R> AIC(casebase_time)
\end{CodeInput}

\begin{CodeOutput}
#> [1] 5821
\end{CodeOutput}

\begin{CodeInput}
R> AIC(casebase_splines)
\end{CodeInput}

\begin{CodeOutput}
#> [1] 5802
\end{CodeOutput}
\end{CodeChunk}

@@

We examined the Study to Understand Prognoses Preferences Outcomes and
Risks of Treatment (SUPPORT) dataset \citeyearpar{knaus1995support}. The
SUPPORT dataset tracks death for individuals who are considered
seriously ill within a hospital. Imputation was conducted using the
default methods in the \pkg{mice} package in \proglang{R}, available on
CRAN.
\textcolor{red}{Why was this method used? The Vanderbilt University website lists describes another method for imputation that we could have used.}
Before imputation, there was a total of 1000 individuals and 35
variables. After imputation, we have 474 observations where 67.5\% of
individuals in the study experience the event of interest. The SUPPORT
dataset will be used to demonstrate regularization within casebase,
while comparing their absolute risk predictions for a new individual. A
description of each variable can be found in
Table\textasciitilde{}\ref{tab:support1} and a breakdown of each
categorical variable in Table\textasciitilde{}\ref{tab:support2}. The
data was downloaded from the main author's website, and it is also
available as part of the \pkg{casebase} package.

@@

Hospital death as a covariate was removed, as this is directly
informative of death. The \pkg{glmnet} package on CRAN was used to
introduce elastic net, lasso and ridge regressions to \pkg{casebase}. To
visualize the effect of imputation on incidence density, we can refer to
the pre-imputation population-time plot in Figure \ref{fig:ptPreI} and
post-imputation population time plot in Figure \ref{fig:ptPostI}. The
incidence density is consistent between imputed and complete datasets.

@@

\begin{CodeInput}
R> popTime_post <- popTime(as.data.frame(workingCompleteData),
R>                         time = "d.time", event = "death")
R> plot(popTime_post)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebaseAbsolute <- absoluteRisk(cb.Model,
R>                                  time = seq(1, max(completeData$d.time), 0.5),
R>                                  newdata = t(newData),
R>                                  s = "lambda.1se", method = "numerical")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(survival)
R> library(casebase)
R> 
R> stanford_popTime <- popTime(jasa, time = "futime", 
R+                             event = "fustat")
R> plot(stanford_popTime)
\end{CodeInput}

\begin{center}\includegraphics{../figures/stanford-poptime-1} \end{center}

@@

\begin{CodeInput}
R> library(tidyverse)
R> library(lubridate)
R> 
R> cb_data <- sampleCaseBase(jasa, time = "futime", 
R+                           event = "fustat", ratio = 10)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Define exposure variable
R> cb_data <- mutate(cb_data,
R+                   txtime = time_length(accept.dt %--% tx.date, 
R+                                        unit = "days"),
R+                   exposure = case_when(
R+                     is.na(txtime) ~ 0L,
R+                     txtime > futime ~ 0L,
R+                     txtime <= futime ~ 1L
R+                   ))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(splines)
R> # Fit several models
R> fit1 <- fitSmoothHazard(fustat ~ exposure,
R+                         data = cb_data, time = "futime")
R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
R+                         data = cb_data, time = "futime")
R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
R+                         data = cb_data, time = "futime")
R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
R+                         data = cb_data, time = "futime")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Compute AIC
R> c("Model1" = AIC(fit1),
R+   "Model2" = AIC(fit2),
R+   "Model3" = AIC(fit3),
R+   "Model4" = AIC(fit4))
\end{CodeInput}

\begin{CodeOutput}
#> Model1 Model2 Model3 Model4 
#>    810    772    741    722
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> hazard_data <- expand.grid(exposure = c(0, 1),
R+                            futime = seq(0, 1000,
R+                                         length.out = 100))
R> # Set the offset to zero
R> hazard_data$offset <- 0 
R> # Use predict to get the fitted values, and exponentiate to 
R> # transform to the right scale
R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
R+                                  type = "link"))
R> # Add labels for plots
R> hazard_data$Status = factor(hazard_data$exposure,
R+                             labels = c("NoTrans", "Trans"))
R> 
R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
R+     geom_line() +
R+     theme_minimal() +
R+     theme(legend.position = 'top') +
R+     ylab('Hazard') + xlab('Follow-up time')
\end{CodeInput}

\begin{center}\includegraphics{../figures/stanford-hazard-1} \end{center}

>>>>

\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \DeclareUnicodeCharacter{2500}{-}

@@

At the time of writing, a cursory examination of CRAN's task view on
survival analysis reveals that there are over 250 packages related to
survival analysis \citeyearpar{survTaskView}. For the purposes of this
article, we restricted our description to packages that implement at
least one of the following features: parametric modeling,
non-proportional hazard models, competing risk analysis, penalized
estimation, and cumulative incidence curve estimation. By searching for
appropriate keywords in the \code{DESCRIPTION} file of these packages,
we found 60 relevant packages. These 60 packages were then manually
examined to determine which ones are comparable to \pkg{casebase}. In
particular, we excluded packages that were focused on a different set of
problems, such as frailty and multi-state models. The remaining 14
packages appear in Table \ref{tab:surv-pkgs}, along with a description
of some of the functionalities they offer.

Several packages implement penalized estimation for the Cox model:
\pkg{glmnet} \citeyearpar{regpathcox}, \pkg{glmpath}
\citeyearpar{park_hastie}, \pkg{penalized} \citeyearpar{l1penal},
\pkg{RiskRegression} \citeyearpar{gerds_blanche}. Moreover, some
packages also include penalized estimation in the context of Cox models
with time-varying coefficients: \pkg{CoxRidge}
\citeyearpar{perperoglou}, \pkg{rstpm2} \citeyearpar{clements_liu} and
\pkg{survival} \citeyearpar{survival-package}. On the other hand,
\pkg{casebase} provides penalized estimation in the context of
parametric hazards. To our knowledge, this is the only package to offer
this functionality.
\textcolor{red}{Is this true? Because it seems to be a logical conclusion from your paragraph.}

@@

\begin{CodeInput}
R> data(ERSPC)
R> ERSPC$ScrArm <- factor(ERSPC$ScrArm, 
R>                        levels = c(0, 1), 
R>                        labels = c("Control group", "Screening group"))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> pt_object <- casebase::popTime(ERSPC, event = "DeadOfPrCa")
R> plot(pt_object)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> pt_object_strat <- casebase::popTime(ERSPC, 
R>                                      event = "DeadOfPrCa", 
R>                                      exposure = "ScrArm")
R> plot(pt_object_strat)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> plot(pt_object_strat, ncol = 2)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> cox_model <- survival::coxph(Surv(Follow.Up.Time, DeadOfPrCa) ~ ScrArm, 
R>                              data = ERSPC)
R> summary(cox_model)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R>                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> legend("topleft", 
R>        legend = c("Control group", "Screening group"), 
R>        col = c("red","blue"),
R>        lty = c(1, 1), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_exponential <- casebase::fitSmoothHazard(DeadOfPrCa ~ ScrArm, 
R>                                                   data = ERSPC, 
R>                                                   ratio = 100)
R> 
R> summary(casebase_exponential)
R> exp(coef(casebase_exponential)[2])
R> exp(confint(casebase_exponential)[2,])
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_exp <- casebase::absoluteRisk(object = casebase_exponential, 
R>                                           time = seq(0,15,0.1), 
R>                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,2], col = "red", lty = 2)
R> lines(smooth_risk_exp[,1], smooth_risk_exp[,3], col = "blue", lty = 2)
R> 
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_time <- fitSmoothHazard(DeadOfPrCa ~ Follow.Up.Time + ScrArm, 
R>                                  data = ERSPC, 
R>                                  ratio = 100)
R> 
R> summary(casebase_time)
R> exp(coef(casebase_time))
R> exp(confint(casebase_time))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_time <- casebase::absoluteRisk(object = casebase_time, 
R>                                           time = seq(0,15,0.1), 
R>                                           newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_time[,1], smooth_risk_time[,2], col = "red", lty = 2)
R> lines(smooth_risk_time[,1], smooth_risk_time[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> casebase_splines <- fitSmoothHazard(DeadOfPrCa ~ bs(Follow.Up.Time) + ScrArm, 
R>                                     data = ERSPC, 
R>                                     ratio = 100)
R> 
R> summary(casebase_splines)
R> exp(coef(casebase_splines))
R> exp(confint(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> smooth_risk_splines <- absoluteRisk(object = casebase_splines, 
R>                                     time = seq(0,15,0.1), 
R>                                     newdata = new_data)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,2], col = "red", lty = 2)
R> lines(smooth_risk_splines[,1], smooth_risk_splines[,3], col = "blue", lty = 2)
R> 
R> legend("topleft", 
R>        legend = c("Control group (Cox)","Control group (Casebase)",
R>                   "Screening group (Cox)", "Screening group (Casebase)"), 
R>        col = c("red","red", "blue","blue"),
R>        lty = c(1, 2, 1, 2), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

@@

\begin{CodeInput}
R> anova(casebase_exponential, casebase_splines, test = "LRT")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> AIC(casebase_exponential)
R> AIC(casebase_time)
R> AIC(casebase_splines)
\end{CodeInput}
\end{CodeChunk}

@@

We examined the Study to Understand Prognoses Preferences Outcomes and
Risks of Treatment (SUPPORT) dataset \citeyearpar{knaus1995support}. The
SUPPORT dataset tracks death for individuals who are considered
seriously ill within a hospital. Imputation was conducted using the
default methods in the \pkg{mice} package in \proglang{R}, available on
CRAN. Before imputation, there was a total of 1000 individuals and 35
variables. After imputation, we have 474 observations where 67.5\% of
individuals in the study experience the event of interest. The SUPPORT
dataset will be used to demonstrate regularization within casebase,
while comparing their absolute risk predictions for a new individual. A
description of each variable can be found in Table \ref{tab:support1}
and a breakdown of each categorical variable in Table
\ref{tab:support2}. The data was downloaded from the main author's
website, and it is also available as part of the \pkg{casebase} package.

@@

Hospital death as a covariate was removed, as this is directly
informative of death. The glmnet package(REF) on CRAN was used to
introduce elastic net, lasso and ridge regressions to casebase. To
visualize the incidence density, we can refer to the post-imputation
population-time plot in Figure \ref{fig:ptPostI}.

@@

\begin{CodeInput}
R> popTime_post <- popTime(as.data.frame(completeData),
R>                         time = "d.time", event = "death")
R> plot(popTime_post)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> 
R> casebaseAbsolute <- absoluteRisk(cb.Model,
R>                                  time = seq(1, max(completeData$d.time), 0.5),
R>                                  newdata = t(newData),
R>                                  s = "lambda.1se", method = "numerical")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(survival)
R> library(casebase)
R> 
R> stanford_popTime <- popTime(jasa, time = "futime", 
R>                             event = "fustat")
R> plot(stanford_popTime)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(tidyverse)
R> library(lubridate)
R> 
R> cb_data <- sampleCaseBase(jasa, time = "futime", 
R>                           event = "fustat", ratio = 10)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Define exposure variable
R> cb_data <- mutate(cb_data,
R>                   txtime = time_length(accept.dt %--% tx.date, 
R>                                        unit = "days"),
R>                   exposure = case_when(
R>                     is.na(txtime) ~ 0L,
R>                     txtime > futime ~ 0L,
R>                     txtime <= futime ~ 1L
R>                   ))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> library(splines)
R> # Fit several models
R> fit1 <- fitSmoothHazard(fustat ~ exposure,
R>                         data = cb_data, time = "futime")
R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
R>                         data = cb_data, time = "futime")
R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
R>                         data = cb_data, time = "futime")
R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
R>                         data = cb_data, time = "futime")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Compute AIC
R> c("Model1" = AIC(fit1),
R>   "Model2" = AIC(fit2),
R>   "Model3" = AIC(fit3),
R>   "Model4" = AIC(fit4))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> hazard_data <- expand.grid(exposure = c(0, 1),
R>                            futime = seq(0, 1000,
R>                                         length.out = 100))
R> # Set the offset to zero
R> hazard_data$offset <- 0 
R> # Use predict to get the fitted values, and exponentiate to 
R> # transform to the right scale
R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
R>                                  type = "link"))
R> # Add labels for plots
R> hazard_data$Status = factor(hazard_data$exposure,
R>                             labels = c("NoTrans", "Trans"))
R> 
R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
R>     geom_line() +
R>     theme_minimal() +
R>     theme(legend.position = 'top') +
R>     ylab('Hazard') + xlab('Follow-up time')
\end{CodeInput}
\end{CodeChunk}
